Episodic memory: mental time travel or a quantum `memory wave' function?

### 2020-09-16 paper/main.tex
added section on matching problem, working on word embedding section

\begin{figure}[tp] \centering \caption{\textbf{Memory wave functions. All panels.} Each panel displays a quantum memory wave function for a hypothetical individual at a particular moment (time $t = 0$).  The $x$-axes correspond to the position ($t$) along an autobiographical timeline.  The $y$-axes display how much the individual's thoughts at time $t = 0$ include information about the individual's past (or future) perceived or predicted experiences at other times. \textbf{A. Contextual drift (no explicit reinstatement).}  The blend of thoughts leading up to $t = 0$ reflects a recency-weighted average of information from the past.  \textbf{B. Mental time travel.} This constitutes a special case of the quantum memory wave function framework, whereby the individual's thoughts reflect a recency-weighted blend of prior thoughts, along with thoughts specific to one remembered prior experience (occuring at the timepoint indicated by the $\uparrow$ symbol).  Each curve's shading varies with the amount of temporal blurring.  In this example, greater amounts of blurring denotes that the individual is bringing to mind information that is farther from the remembered moment or from $t = 0$.  \textbf{C. Quantum memory waves.}  In this example, the individual remembering the moment indicated by the $\uparrow$ symbol brings back a blend of thoughts similar to those shown in Panel B.  In addition, they bring to mind information pertaining to other conceptually related (but temporally non-contiguous) moments, indicated by $\star$ symbols. As in Panel B, different curves are shaded according to their degree of temporal blurring.  \textbf{D. Generating predictions about the future.} To predict what will happen during a future event ($\diamond$) the individual spreads their thoughts over their prior experiences according to the expected relevance of those experiences to the future event.  Local peaks in the curve denote prior experiences with high expected relevance.  The black part of the curve illustrates how the individual's thoughts are spread out over their past, and the gray part of the curve illustrates how the individual's thoughts are spread out over their (predicted) future.}
\label{fig:qwave}
\end{figure}

@@

One insight into this distinction comes from a recent study by \cite{HeusEtal18c}, which analyzed data collected by \cite{ChenEtal17} who had participants watch an episode of the BBC television show \textit{Sherlock} and then verbally recount what happened in the episode.  The authors used topic models~\citep{BleiEtal03} applied to human-generated annotations of each camera shot in the episode to characterize the episode's content with a timeseries of high-dimensional semantic feature vectors.  These \textit{topic vectors} describe the mix of ``themes'' present in each moment of the episode.  When characterized in this way, the episode's content exhibits periods of relative stability, followed by moments of rapid change. This interplay between stability and change is visible as ``blocks'' in the temporal correlation matrix (Fig.~\ref{fig:corrmats}A); \cite{HeusEtal18c} performed detailed analyses of these putative \textit{events} and how they are remembered.  Notably, the off-diagonal entries of the temporal correlation matrix (which reflect content similarity between temporally distant events in the episode) are nearly all close to 0.  This is surprising, because the episode itself comprises a murder mystery with recurring motifs that must be associated in the viewer's mind in order to make sense of the episode.  Nonetheless, when considering solely the semantic content of each moment of the episode, those long timescale associations are nearly entirely absent.  To the extent that this finding extends to real-world experiences, it comports with Heraclitus' notion that one cannot step in the same river twice~\citep{Hera}.

\begin{figure}[tp] \centering \includegraphics[width=0.8\textwidth]{figs/pres_rec_corrmats_sherlock.png} \caption{\textbf{Temporal correlation matrices of the dynamic content of a television episode and people's recalls of that episode.}  The panels of this figure are adapted from \cite{HeusEtal18c}, where full details may be found. \textbf{A. Temporal correlations exhibited by the content of the television episode.}  Each moment of an episode of the BBC television show \textit{Sherlock} has been characterized using a topic model~\citep{BleiEtal03} fit to detailed human-generated annotations between each scene cut.  The correlations between the topic vectors of a given pair of timepoints reflects the similarity in thematic content at those timepoints.  \textbf{B. Temporal correlations exhibited by the (average) content of participants' recalls of the episode.} The topic model used to characterize the episode in Panel A has been applied to each sentence from each participants' recall transcript.  The temporal correlation matrices describing each participants' recalls were re-sampled using cubic interpolation to have 100 rows and columns prior to averaging the resampled matrices across participants.  \textbf{C. Temporal correlations in the content of individual participants' recalls of the episode.} Each sub-panel displays the temporal correlation matrix for one experimental participant.}
\label{fig:corrmats}
\end{figure}

@@

To understand which specific events are described (during recall) as more similar than what their original experiences (as characterized by their topic vectors) would suggest, it can be helpful to examine individual recalls of particular events throughout the episode.  Figure~\ref{fig:reinstatement}B displays, for a representative example recalled moment, which other moments from the episode were characterized in a similar way by the participants when they recounted the episode later (black and colored lines), and which other moments in the episode were described in a similar way in moment-by-moment (non-recall) human annotations (gray line).  The example event explored in the figure comprises a scene where Sherlock Holmes and John Watson (main characters) are bantering, and then are interrupted when Sherlock realizes that another victim has been murdered.  The language used by human annotators describing the moment-by-moment contents of the scene used a similar mix of themes only for moments in the episode that were temporally proximal to the event in question (gray line).  However, the way participants described what happened when they \textit{recounted} the episode later revealed a much different pattern.  Specifically, their language mirrored that used when they were recalling other murders, and other moments when Sherlock and John were bantering with or about each other (Fig.~\ref{fig:reinstatement}C).  The way they described the example event did \textit{not} mirror the way they described other semantically (but not conceptually) related events, such as events involving unrelated violence, or Sherlock and John interacting with other characters about other subjects.  Taken together, this suggests that these specific events were associated in participants' memories, perhaps contributing to (or reflecting) their understanding that those events were linked in the episode's narrative.

\begin{figure}[tp] \centering \includegraphics[width=\textwidth]{figs/reinstatement.png} \caption{\textbf{Content drift in a television episode and participants' recalls of that episode.  A.  Autocorrelations in the episode's content and in the content of participants' recalls.}  The black line denotes the average correlations between the topic vectors from pairs of moments in the episode as a function of their temporal separation (times are relative to the full length of the episode).  Each colored line denotes an analogous autocorrelation function, but for each individual participant's recalls of the episode (times are relative to the full length of each participant's recalls).  The error ribbons denote 95\% confidence intervals, taken across all moments in the episode (or in participants' recalls of the episode).  \textbf{B. Recall density functions for one recalled moment.} Each curve reflects the correlations between the topic vector at each (relative) moment of the episode (or participants' recalls of the episode), and the topic vector of an event beginning approximately $\frac{1}{3}$ through the episode (relative time: 0.328).  The gray curve denotes these correlations for the topic vectors derived from annotations of the episode itself. The black line denotes the average correlations across all participants' recalls.  Local maxima are marked with colored dots.  Local maxima closer than 2.5\% of the total recall duration to a higher local maximum were excluded (this constraint served to eliminate very nearby peaks that corresponded to duplicate events), as were maxima with peaks with correlation coefficients lower than 0.1 (this threshold was chosen arbitrarily; this constraint served to eliminate events that were only very weakly associated with the example event).   The colored lines denote the correlations for each individual participant's recalls.  \textbf{C. Local maxima in the average recall density function.}  Representative screen captures from the scenes at each local maxima in Panel B are displayed (dot colors match those in Panel B).  The text in each screen capture displays a brief summary of what happened in each scene. The colored outline denotes the reference scene being recalled.}
  % colors: individual participants
  % black: average across participants
  % gray: autocorrelation with topic vector at timepoint
  % circles: local maxima in the across-participants average (i.e., movie events that share structure with the given timepoint that is distinct from other temporally proximal events)
\label{fig:reinstatement}
\end{figure}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.8\textwidth]{figs/trajectory.png}
\caption{\textbf{The shape of an unfolding experience.}  \textbf{A. The trajectory through topic space taken by a television episode.} The topic trajectory taken by a 50 minute television episode, projected onto two dimensions.  Each dot indicates an event identified using a Hidden Markov Model; the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B. The trajectory through topic space taken by participants' recalls of the television episode.} The average two-dimensional trajectory captured by 17 participants' recalls of the episode, with the same format and coloring as the trajectory in Panel A. The arrows reflect the directions of the average transitions taken by participants whose trajectories crossed that part of topic space; blue denotes reliable agreement across participants.  Figure adapted from \cite{HeusEtal18c}, where additional details may be found.}
\label{fig:trajectories}
\end{figure}

In addition to serving as a framework for quantifying the dynamic content of experiences, these trajectories may be visualized geometrically by projecting word embedding representations of the events onto lower dimensional spaces~\citep[e.g., ][]{HeusEtal18a}; for additional discussion and counterpoints see~\cite{JollChan18}.  \cite{HeusEtal18c} applied this approach to the same dataset reflected in Figure~\ref{fig:trajectories}, revealing another key property of how we remember.  Although there are substantial individual differences in the specific words people use to describe their shared experience of watching the same television episode, and how much detail they recollected, the overall shapes of those recalled experience trajectories are highly similar across people.  This has potentially exciting implications for what it means to communicate memories to other people: we need to convey the ``gist'' (shape) of our experience.  It also has implications for how we learn by analogy: perhaps events that share schemas~\citep[e.g., ][]{BaldEtal18}  unfold in similar ways (e.g., have similar shapes).  Learning which shapes go with which schemas could provide a scaffolding for rapid learning and generalization of new experiences, to the extent that those new experiences match a previously learned schema (shape).

>>>>

\begin{figure}[tp] \centering \includegraphics[width=\textwidth]{figs/wave_functions}
  \caption{\textbf{Memory wave functions. All panels.} Each panel displays a quantum memory wave function for a hypothetical individual at a particular moment (time $t=0$ indicated by $\blacktriangle$)).  The $x$-axes correspond to the position ($t$) along an autobiographical timeline.  The $y$-axes display how much the individual's thoughts at time $t = 0$ include information about the individual's past (or future) perceived or predicted experiences at other times. All curves have been normalized to have a maximum value of 1 and a minimum value of 0.  \textbf{A. Contextual drift (no explicit reinstatement).}  The blend of thoughts leading up to $t = 0$ reflects a recency-weighted average of information from the past.  \textbf{B. Mental time travel.} This constitutes a special case of the quantum memory wave function framework, whereby the individual's thoughts reflect a recency-weighted blend of prior thoughts, along with thoughts specific to one remembered prior experience (occuring at the timepoint indicated by the $\bigstar$ symbol).  Each curve's coloring denotes the amount of temporal blurring.  In this example, greater amounts of blurring denotes that the individual is bringing to mind information that is farther from the remembered moment or from $t = 0$.  \textbf{C. Quantum memory waves.}  In this example, the individual remembering the moment indicated by the $\bigstar$ symbol brings back a blend of thoughts similar to those shown in Panel B.  In addition, they bring to mind information pertaining to other conceptually related (but temporally non-contiguous) moments, indicated by $\star$ symbols. As in Panel B, different curves are colored according to their degree of temporal blurring.  \textbf{D. Generating predictions about the future.} To predict what will happen during a series of future events ($\diamond$ symbols) the individual spreads their thoughts over their prior experiences according to the expected relevance of those experiences to the future events.  Local peaks in the curve denote prior experiences with high expected relevance.  The left part of the panel (prior to $\blacktriangle$) illustrates how the individual's thoughts are spread out over their past, and the right part of the panel illustrates how the individual's thoughts are spread out over their (predicted) future.}
\label{fig:qwave}
\end{figure}

@@

One insight into this distinction comes from a recent study by \cite{HeusEtal18c}, which analyzed data collected by \cite{ChenEtal17} who had participants watch an episode of the BBC television show \textit{Sherlock} and then verbally recount what happened in the episode.  The authors used topic models~\citep{BleiEtal03} applied to human-generated annotations of each camera shot in the episode to characterize the episode's content with a timeseries of high-dimensional semantic feature vectors.  Each shot's annotation contained the following information: a brief narrative description of what was happening, the location where the scene took place, whether that location was indoors or outdoors, the names of all on-screen characters, the names of the character(s) who were in focus in the camera shot, the name(s) of any character(s) who spoke during the scene, the camera angle of the shot, a transcription of any on-screen text, and whether or not there was any music present in the auditory background~\citep[full details may be found in][]{ChenEtal17}.  The resulting \textit{topic vectors} describe the mix of ``themes'' present in each moment of the episode.  When characterized in this way, the episode's content exhibits periods of relative stability, followed by moments of rapid change. This interplay between stability and change is visible as ``blocks'' in the temporal correlation matrix (Fig.~\ref{fig:corrmats}A); \cite{HeusEtal18c} performed detailed analyses of these putative \textit{events} and how they are remembered.  Notably, the off-diagonal entries of the temporal correlation matrix (which reflect content similarity between temporally distant events in the episode) are nearly all close to 0.  This is surprising, because the episode itself comprises a murder mystery with recurring motifs that must be associated in the viewer's mind in order to make sense of the episode.  Nonetheless, when considering solely the semantic content of each moment of the episode, those long timescale associations are nearly entirely absent.  To the extent that this finding extends to real-world experiences, it comports with Heraclitus' notion that one cannot step in the same river twice~\citep{Hera}.

\begin{figure}[tp] \centering \includegraphics[width=0.8\textwidth]{figs/pres_rec_corrmats_sherlock} \caption{\textbf{Temporal correlation matrices of the dynamic content of a television episode and people's recalls of that episode.}  The panels of this figure are adapted from \cite{HeusEtal18c}, where full details may be found. \textbf{A. Temporal correlations exhibited by the content of the television episode.}  Each moment of an episode of the BBC television show \textit{Sherlock} has been characterized using a topic model~\citep{BleiEtal03} fit to detailed human-generated annotations between each scene cut.  The correlations between the topic vectors of a given pair of timepoints reflects the similarity in thematic content at those timepoints.  \textbf{B. Temporal correlations exhibited by the (average) content of participants' recalls of the episode.} The topic model used to characterize the episode in Panel A has been applied to each sentence from each participants' recall transcript.  The temporal correlation matrices describing each participants' recalls were re-sampled using cubic interpolation to have 100 rows and columns prior to averaging the resampled matrices across participants.  \textbf{C. Temporal correlations in the content of individual participants' recalls of the episode.} Each sub-panel displays the temporal correlation matrix for one experimental participant.}
\label{fig:corrmats}
\end{figure}

@@

To understand which specific events are described (during recall) as more similar than what their original experiences (as characterized by their topic vectors) would suggest, it can be helpful to examine individual recalls of particular events throughout the episode.  Figure~\ref{fig:reinstatement}B displays, for a representative example recalled moment, which other moments from the episode were characterized in a similar way by the participants when they recounted the episode later (black and colored lines), and which other moments in the episode were described in a similar way in moment-by-moment (non-recall) human annotations (gray line).  The example event explored in the figure comprises a scene where Sherlock Holmes and John Watson (main characters) are bantering, and then are interrupted when Sherlock realizes that another victim has been murdered.  The language used by human annotators describing the moment-by-moment contents of the scene used a similar mix of themes only for moments in the episode that were temporally proximal to the event in question (gray line).  However, the way participants described what happened when they \textit{recounted} the episode later revealed a much different pattern.  Specifically, their language mirrored that used when they were recalling other murders, and other moments when Sherlock and John were bantering with or about each other (Fig.~\ref{fig:reinstatement}C).

The way they described the example event did \textit{not} mirror the way they described other semantically (but not conceptually) related events, such as events involving unrelated violence, or Sherlock and John interacting with other characters about other subjects.  Taken together, this suggests that these specific events were associated in participants' memories, perhaps contributing to (or reflecting) their understanding that those events were linked in the episode's narrative.

One concrete example of this phenomenon is that, early in the episode, Sherlock and John appear to be investigating a series of suicides.  Later in the episode, it becomes clear that the apparent suicides are actually murders.  The annotations refer to early deaths as suicides, whereas participants often refer to those early deaths as ``murders'' even though they could not have known that they were murders early on in the episode.  In this way, participants' experiences \textit{subsequent to} a target remembered event may distort its memory by incorporating information learned during other experiences.

\begin{figure}[tp] \centering \includegraphics[width=\textwidth]{figs/reinstatement} \caption{\textbf{Content drift in a television episode and participants' recalls of that episode.  A.  Autocorrelations in the episode's content and in the content of participants' recalls.}  The black line denotes the average correlations between the topic vectors from pairs of moments in the episode as a function of their temporal separation (times are relative to the full length of the episode).  Each colored line denotes an analogous autocorrelation function, but for each individual participant's recalls of the episode (times are relative to the full length of each participant's recalls).  The error ribbons denote 95\% confidence intervals, taken across all moments in the episode (or in participants' recalls of the episode).  \textbf{B. Recall density functions for one recalled moment.} Each curve reflects the correlations between the topic vector at each (relative) moment of the episode (or participants' recalls of the episode), and the topic vector of an event beginning approximately $\frac{1}{3}$ through the episode (relative time: 0.328).  The gray curve denotes these correlations for the topic vectors derived from annotations of the episode itself. The black line denotes the average correlations across all participants' recalls.  Local maxima are marked with colored dots.  Local maxima closer than 2.5\% of the total recall duration to a higher local maximum were excluded (this constraint served to eliminate very nearby peaks that corresponded to duplicate events), as were maxima with peaks with correlation coefficients lower than 0.1 (this threshold was chosen arbitrarily; this constraint served to eliminate events that were only very weakly associated with the example event).   The colored lines denote the correlations for each individual participant's recalls.  \textbf{C. Local maxima in the average recall density function.}  Representative screen captures from the scenes at each local maxima in Panel B are displayed (dot colors match those in Panel B).  The text in each screen capture displays a brief summary of what happened in each scene. The colored outline denotes the reference scene being recalled.}
  % colors: individual participants
  % black: average across participants
  % gray: autocorrelation with topic vector at timepoint
  % circles: local maxima in the across-participants average (i.e., movie events that share structure with the given timepoint that is distinct from other temporally proximal events)
\label{fig:reinstatement}
\end{figure}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.8\textwidth]{figs/trajectory}
\caption{\textbf{The shape of an unfolding experience.}  \textbf{A. The trajectory through topic space taken by a television episode.} The topic trajectory taken by a 50 minute television episode, projected onto two dimensions.  Each dot indicates an event identified using a Hidden Markov Model; the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B. The trajectory through topic space taken by participants' recalls of the television episode.} The average two-dimensional trajectory captured by 17 participants' recalls of the episode, with the same format and coloring as the trajectory in Panel A. The arrows reflect the directions of the average transitions taken by participants whose trajectories crossed that part of topic space; blue denotes reliable agreement across participants.  Figure adapted from \cite{HeusEtal18c}, where additional details may be found.}
\label{fig:trajectories}
\end{figure}

In addition to serving as a framework for quantifying the dynamic content of experiences, these trajectories may be visualized geometrically by projecting word embedding representations of the events onto lower dimensional spaces~\citep[e.g., ][]{HeusEtal18a}; for additional discussion and counterpoints see~\cite{JollChan18}.  \cite{HeusEtal18c} applied this approach to the same dataset reflected in Figure~\ref{fig:trajectories}, revealing another key property of how we remember.  Although there are substantial individual differences in the specific words people use to describe their shared experience of watching the same television episode, and how much detail they recollected, the overall shapes of those recalled experience trajectories are highly similar across people.  This has potentially exciting implications for what it means to communicate memories to other people: we need to convey the ``essence'' (shape) of our experience.  It also has implications for how we learn by analogy: perhaps events that share schemas~\citep[e.g., ][]{BaldEtal18}  unfold in similar ways (e.g., have similar shapes).  Learning which shapes go with which schemas could provide a scaffolding for rapid learning and generalization of new experiences, to the extent that those new experiences match a previously learned schema (shape).

\subsection*{The \textit{matching problem}}
The trajectories through word embedding space displayed in Figure~\ref{fig:trajectories} demonstrate that topic models are sufficiently expressive and robust to provide one solution to the \textit{matching problem} in memory analysis.  The matching problem refers to discovering an alignment in the labels between experienced and remembered events.  Traditional memory paradigms most commonly employed over the past century~\citep{Kaha12} ask participants to study lists of \textit{items} (often randomly selected words) and then engage in a memory task, such as verbally recalling any words the participant remembers studying, in any order.  The matching problem in such experiments is trivial: if we treat each studied word on a list as an ``event,'' then we can assume that when the participant verbalizes that word during their memory test, they are remembering that word's event.  In other words, matching entails finding the exact match between a word the participant is verbalizing and that same word on the list that they studied.

For real-world memory phenomena, and in many so-called \textit{naturalistic} memory paradigms, solving the matching problem is more complex.  People often experience real-world events (and scenes from movies or stories) as rich sensory experiences that derive deep meaning from surrounding moments, other prior experiences, emotions, goals, high-order understandings of the current situation, and so on.  Many of these aspects of our real-world experiences are not immediately verbalizable-- and even when we are able to describe some specific aspects of what we are experiencing, there are typically many possibly ways we could have done so (e.g., varying in level of detail, specific wording, objectivity versus subjectivity, contextual details, etc.).  Further, because the way we experience each moment depends in part on our unique prior experiences, even ostensibly shared experiences can seem different to each individual.  Similarly, the way we recount a given experience is often a reflection of our own idiosyncratic experience of the original event, colored and distorted in an idiosyncratic way by our memory systems.  This distorted memory must then be verbalized or communicated, which is itself an idiosyncratic (and potentially lossy in the information theoretic sense) process.  Solving the matching problem under these constraints requires building explicit models of the dynamic conceptual content of our experiences and memories, and then estimating when and how that experienced and remembered content aligns.  The approach featured in Figure~\ref{fig:trajectories} uses a word embedding space that is defined by a topic model~\citep{BleiEtal03}.  Each coordinate in this space reflects the conceptual content underlying a set of words, such that two different text passages that describe the same conceptual content (but using different words, levels of detail, etc.) are mapped on to nearby points in the word embedding space.  When these models are sufficiently expressive and robust to model our experiences and memories within the same word embedding space, we can see a clear geometric alignment between the trajectories that our experiences (Fig.~\ref{fig:trajectories}A) and memories (Fig.~\ref{fig:trajectories}B) take through word embedding space~\citep[also see][]{HeusEtal18c}.

\subsection*{Defining the geometries of ongoing thoughts and memories}
%JRM STOPPED HERE
